\documentclass[a4paper]{report}
%\usepackage{afterpage}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{caption}
\usepackage[mathscr]{eucal}
\usepackage{graphicx}
\usepackage{hyperref}
\usepackage{pgfplots}
\usepackage{subfigure}
\usepackage{xcolor}

\include{etc/commands}

\title{\texttt{bnplib}: A Nonparametric C++ Library}

\author{
Bruno Guindani \\
Elena Zazzetti \\
\\
Professors: Alessandra Guglielmi, Luca Formaggia \\
Advisor: Mario Beraha \\
\\
\\
Politecnico di Milano \\
\\
}

\date{
June 12th, 2020
\\[250pt]
{\color{gray} {\url{https://github.com/poliprojects/BNPlib}}}
}

\begin{document}

\maketitle

\newpage\phantom{a}

\thispagestyle{empty}
\begin{abstract}
We present a C++ library that exploits a Bayesian Non-Parametric setting in order to conduct unidimensional and multidimensional data analysis.
In such a setting, our main goals are density estimation and clustering analysis.
Several iterative algorithms are available that make use of Gibbs sampling, building a Markov chain that reaches convergence at a reasonably fast pace.
In particular, we focused on implementing some algorithms introduced by Neal in 2000.
After running one of these algorithms, density and cluster estimation can then be conducted by using the provided auxiliary tools.
In this report, after an overview of the underlying Bayesian model and a roundup of the algorithms' state of the art, we delve into the details of our implementation and then present an example of data analysis, whilst providing the theoretical background for our estimates.
We also include a description for the implemented Python interface for this C++ library.
\end{abstract}
\setcounter{page}{4}

\tableofcontents

% Part 1
\include{01_intro}
\include{02_algorithms_1}
\include{03_estimates_1}

% Part 2
\include{04_hierarchies}
% TODO why is there a page break here? fix
\include{05_algorithms_2}
\include{06_collectors}
\include{07_estimates_2}
\include{08_factory}
\include{09_python}
\include{10_performance}
\include{11_sensit_an}
\include{12_tests_py}
\include{13_extensions}


\begin{thebibliography}{9}
	\bibitem{book} P. Muller, F. A. Quintana, \textit{Bayesian Nonparametric Data Analysis}

	\bibitem{neal} R. M. Neal (2000), \textit{Markov Chain Sampling Methods for Dirichlet Process Mixture Models}

	\bibitem{james} H. Ishwaran, L. F. James (2001), \textit{Gibbs Sampling Methods for Stick-Breaking Priors}

	\bibitem{PY} J. Pitman, M. Yor (1997), \textit{The two-parameter Poisson-Dirichlet distribution derived from a stable subordinator} 

	\bibitem{integral} K. P. Murphy (2007), \textit{Conjugate Bayesian analysis of the Gaussian distribution}

	\bibitem{rand} Rand W. (1971), \textit{Objective Criteria for the Evaluation of Clustering Methods.
	Journal of the American Statistical Association}, 66(336), 846-850. doi:10.2307/2284239

	\bibitem{stan} Stan documentation: \url{http://mc-stan.org/math}. \\
	Code found at \url{https://github.com/stan-dev/math} (current version 3.2.0 was used) and it also includes other needed libraries: Boost, Eigen, SUNDIALS, Intel TBB

	\bibitem{eigen} Eigen documentation: \url{https://eigen.tuxfamily.org/dox}. \\
	Code is included in the Stan package (current version 3.3.7 was used)

	\bibitem{proto} Protocol Buffers Tutorial for C++: \url{https://developers.google.com/protocol-buffers/docs/cpptutorial}. \\
	Code found at \url{https://github.com/protocolbuffers/protobuf} (version 3.11.0 was used)

	\bibitem{pybind11} pybind11 tutorial and documentation: \url{http://pybind11.readthedocs.org/en/master}
	
	\bibitem{skl} shikit-learn user guide: \url{https://scikit-learn.org/stable/user_guide.html}

	\bibitem{tutors} Codes of Mario Beraha and Riccardo Corradin for similar projects, found at \url{https://github.com/mberaha/partial_exchangeability} and \url{https://github.com/rcorradin/BNPmix} respectively

	\bibitem{beep-bayes} Course material for Bayesian Statistics by Prof. Alessandra Guglielmi: \url{https://beep.metid.polimi.it/web/2019-20-bayesian-statistics-alessandra-guglielmi-/}
	
	\bibitem{beep-pacs} Course material for Advanced Programming for Scientific Computing by Prof. Luca Formaggia:
	\url{https://beep.metid.polimi.it/web/102725282}
\end{thebibliography}

\end{document}

%TODO:
% cleanup of bnpmix.R
% rerun per vedere se il fix a Neal8::set_n_aux() è corretto
